---
title: "re_tidy"
author: "Jonathan Lanz 23-106-255"
output: html_document
date: "2025-03-03"
---

In this report, I want to investigate if the CO<sub>2</sub> air content has an effect on soil organic matter.

```{r setup, warning=FALSE, message=FALSE}
# load tidyverse
library(tidyverse)

# load data
soil_carbon_storage <- read_delim("../data/bodendaten_CLEAN.csv", delim = ";")
```

The CSV file I am working with contains Information about the amount of CO<sub>2</sub> depending on the CO<sub>2</sub> content in the air. One measure was made when CO<sub>2</sub> air concentration was ambient, and one when the CO<sub>2</sub> air concentration is elevated. This experiment has been executed at different places and times, and has been running for various amounts of time. The original dataset was downloaded here: 
https://www.science.org/doi/10.1126/science.1249534.
Since the data was not tidy, I manually cleaned it for this exercise. 

My working hypothesis is that the increase of CO<sub>2</sub> in the atmosphere leads to an increase of C in the soil and therefore to an increased soil organic matter. To do so, I define three phases of the experiment and calculate the log-ratio-response (LRR). The LRR indicates whether the change of soil organic matter has a positive or negative trend.

```{r, warning=FALSE}
# define early, mid and late phase
early_phase <- soil_carbon_storage |> 
  filter(Time_years < 3)
mid_phase <- soil_carbon_storage |> 
  filter(Time_years >= 3 & Time_years <= 6)
late_phase <- soil_carbon_storage |>
  filter(Time_years > 6)

# define a function that calculates the LRR
lrr_calculation <-  function(ambient, elevated){
  lrr <-  log(elevated / ambient)
  return(lrr)
}

# calculate log-response ratio (LRR) for every phase
early_phase$lrr <-  lrr_calculation(
  early_phase$ambient_CO2_mean, early_phase$increased_CO2_mean)

mid_phase$lrr <-  lrr_calculation(
  mid_phase$ambient_CO2_mean, mid_phase$increased_CO2_mean)

late_phase$lrr <-  lrr_calculation(
  late_phase$ambient_CO2_mean,late_phase$increased_CO2_mean)

# calculate mean of lrr for every phase
early_phase_lrr_mean <- mean(early_phase$lrr)
mid_phase_lrr_mean <-  mean(mid_phase$lrr)
late_phase_lrr_mean <-  mean(late_phase$lrr)

# Save it in a data frame
lrr_summary <- data.frame(
  phase = c("early", "mid", "late"),
  LRR = c(early_phase_lrr_mean, mid_phase_lrr_mean, late_phase_lrr_mean)
)

knitr::kable(lrr_summary, caption="LRR over time")
```

 
Since the LRR mean is in all three phases close enough to zero (± 0.05) to make it negligible, the results do not provide evidence for my working hypothesis. I conclude that the data available leads to the assumption that the soil organic matter is not influenced by the CO<sub>2</sub> in the air.








